---
title: "Betwenness Centrality"
execute: 
  eval: true
  echo: true
  output: true
  warning: false
  message: false
format: 
   html:
      code-line-numbers: true
---

Recall that, according to @freeman79, the key conceptual distinction between [closeness](closcent.qmd) and betweenness is between (pun intended) the capacity to reach others quickly (e.g., via the shortest paths) and the capacity to *intermediate* among those same paths. High betweenness nodes control the flow of information in the network between other nodes that flows via shortest.

Let us begin by defining some basic concepts on which betweenness centrality measures are built.

### Pairwise Dependency and Betweenness

We can define the **pair dependency** $\delta_{i(k)j}$ [@brandes01, p. 166] of any two nodes in the graph $\{i, j\}$ on any third node $k$ as follows:

$$
\delta_{i(k)j}= \frac{\sigma_{i(k)j}}{\sigma_{ij}}
$$ {#eq-pair}

Where the denominator of the fraction ($\sigma_{ij}$) is a count of the *total* number of shortest paths that start and end with nodes $i$ and $j$ and the numerator of the fraction ($\sigma_{i(k)j}$) is the subset of those paths that include node $k$ as an **inner node**.^[As @brandes08 [p. 137] notes, when $i = j$, then $\sigma_{ij} = 1$ and when $k \in \{i, j\}$, then $\sigma_{i(k)j} = 0$.]

Because @eq-pair is a ratio it can range from zero to one  with everything in between. As such the pair dependency of a pair of nodes relative to a third has an intuitive interpretation as a *probability*, namely the probability that if you send something starting from $i$ in order for it to get to $j$ it has to go *through* $k$. This probability is 1.0 if $k$ stands in *every* shortest path between $i$ and $j$ and zero if $k$ does not lie in *any* of the shortest paths linking $i$ and $j$.

We can also defined the **one-sided dependency** of any one node $i$ on any other node $k$ in the graph as the sum of the pairwise dependencies on $k$ that involve $i$ and every node $j$:
$$
\delta_{i|k} = \sum_j \delta_{i(k)j}
$$ {#eq-one}

The **betweenness centrality** of a node $k$ is then defined as the *sum of the dependencies* that *every pair* of nodes in the graph has on that node:

$$
C^{BET}_k = \sum_{i,j} \frac{\sigma_{i(k)j}}{\sigma_{ij}}
$$ {#eq-bet1}

## Computing Betweenness

Let's see an example of how to calculate betweenness centrality using real data. We first load our trusty *Pulp Fiction* data set from the `networkdata` package, which is an **undirected** graph of character scene co-appearances in the film:

```{r}
    library(networkdata)
    library(igraph)
    library(stringr) #using stringr to change names from all caps to title case
    g <- movie_559
    V(g)$name <- str_to_title(V(g)$name)
    V(g)$name[which(V(g)$name == "Esmarelda")] <- "Esmeralda" #fixing misspelled name
    E(g)$weight <- 1 #setting weights to 1.0 (relevant for betweenness calculation)
    V(g)$name
```

To compute the components of the betweenness centrality defined earlier, we need two pieces of information. First we need to know the *length* of the shortest path between every pair of nodes. As we saw in our discussion of [closeness](closcent.qmd), this is stored in the **geodesic distance matrix** of the graph $\mathbf{G}$. In `igraph` we can obtain that using the `distances` function:

```{r}
    G <- distances(g)
    G[1:10, 1:10]
```

Next we need to know the *number* of shortest paths linking every pair of nodes (as nodes can be linked by multiple distinct shortest paths). In `igraph` we can use the `all_shortest_paths` function to figure out this information. 

For instance, let's say we wanted to find the number of shortest paths between the pair of nodes "Young Man" (at the diner) and "The Pedestrian" (who appears when Butch encounters Marsellus on the street). To do that we can just type:

```{r}
    ap <- all_shortest_paths(g, from = "Pedestrian", to = "Young Man")
    ap$vpaths
```

The `all_shortest_paths` function has three arguments: The graph `g`, the `from` node (the starting node in the path count) and the `to` node (the end node in the path count).^[In an undirected graph you get the same result for a pair of nodes regardless of which one you put on the `from` or `to` slots.] 

The function stores the results in a list object containing various sub-objects. The actual paths (containing the node ids) are in a list sub-object called `vpaths`, which is itself a list (with named vectors containing the indices of the nodes of each path as elements). The `length` of this list is the number of shortest paths between the young man at the diner and the pedestrian:

```{r}
    length(ap$vpaths)
```

We can see that the list is of length eight, indicating that there are eight total shortest paths between the pedestrian and the young man.

If we wanted to verify the *length of the shortest path* between these two nodes, we could just type:

```{r}
    G["Young Man", "Pedestrian"]
```

Which provides us with the relevant entry in the geodesic distance matrix, telling us that the young man is four steps away from the pedestrian in the network. This also means that each path will involve five nodes (counting the young man and the pedestrian, with three intermediary nodes).

Note that we could have gotten the same information using the `distances` function in `igraph` specifying the `v` and `to` nodes:

```{r}
    distances(g, v = "Young Man", to = "Pedestrian")
```

Now, to create the **shortest paths matrix** for the graph $\mathbf{S}$, we can just loop through each pair and record the length of the shortest paths for non-adjancent nodes:

```{r}
    n <- vcount(g)
    A <- as.matrix(as_adjacency_matrix(g))
    S <- matrix(1, n, n) #intializing shortest paths matrix
    diag(S) <- 0 #setting diagonals to zero (Brandes, 2008, p. 137)
    rownames(S) <- V(g)$name
    colnames(S) <- V(g)$name
    for (i in 1:n) {
        for (j in 1:n) {
            if (A[i,j] == 0 & i < j) { #restrict to non-adjancent nodes in upper triangle
                S[i,j] <- length(all_shortest_paths(g, from = i, to = j)$vpaths)
                }
            }
        }
    S[lower.tri(S)] <- t(S)[lower.tri(S)] #copying upper triangle into lower triangle
    S[1:10, 1:10]
```

Once we have this information, we have all we need to compute the various quantities defined earlier. 

For instance, @brandes01 [p. 166] shows that we can take advantage of the fact that a node $k$ can only lie on a shortest path between $i$ and $j$ if the sum of the geodesic distances between $i$ and $k$ and between $k$ and $j$ is equal to the geodesic distance between $i$ and $j$:
$$
g_{ij} = g_{ik} + g_{kj}
$$

If the above equality obtains, then the number of shortest paths between $i$ and $j$ (the numerator of @eq-pair) is simply $s_{ik} \times s_{kj}$, namely the product of the number of shortest paths between $i$ and $k$ and between $k$ and $j$, meaning that the pair dependency of $i$ and $j$ on $k$ will be just:

$$
\delta_{i(k)j}= \frac{s_{ik} s_{kj}}{s_{ij}}
$$

This also means that if the above equality does *not* obtain and $g_{ij} < g_{ik} + g_{kj}$, then we know for sure that the pair dependency of $i$ and $j$ on $k$ is zero ($k$ does not lie on any shortest path between $i$ and $j$). 

Using this approach, we can write a simple function to compute the pair dependency of any two nodes on a third node, using information stored in the $\mathbf{G}$ and $\mathbf{S}$ matrices:

```{r}
    pair.dep <- function(a, b, c) {
        pd <- 0
        if (G[a, b] == G[a, c] + G[c, b]) {
            pd <- (S[a, c] * S[c, b])/S[a, b]
            }
        return(pd)
        }
```

Testing it out to compute the pair dependency of the young man and the pedestrian on Vincent:

```{r}
    pair.dep("Pedestrian", "Young Man", "Vincent")
```

Which says that Vicent lies on `r round(pair.dep("Pedestrian", "Young Man", "Vincent"), 2) * 100`% of the paths between the The Young Man and The Pedestrian (four out of eight).


```{r}
    pair.dep2 <- function(a, b, c) {
        p <- all_shortest_paths(g, from = a, to = b)$vpaths #all shortest paths from i to j
        return(sum(sapply(p, function(x) {c %in% names(x)}))/length(p)) #paths that involve Vincent divided by total
        }             
```

```{r}
    pair.dep("Esmeralda", "Young Man", "Jules")
```


As noted in @eq-bet1, Vincent's *betweenness* is just the sum of the dependencies of each pair of nodes in the graph on him. We could compute that (not very efficiently) using the following wrapper function:

```{r}
    bet.cent <- function(v) {
        bc <- 0
        k <- which(V(g)$name == v)
        node.vec <- c(1:vcount(g))[-k]
        for (i in node.vec) {
            for (j in node.vec) {
                if (i < j & are_adjacent(g, i, j) == FALSE) {
                    bc <- bc + pair.dep(i, j, k)
                }
            }
        }
    return(bc)
    }
    bet.cent("Vincent")
```
