---
title: "Using Edge Clustering to Assign Nodes to Overlapping Communities "
execute: 
  eval: true
  echo: true
  output: true
  warning: false
  message: false
format: 
   html:
      code-line-numbers: true
---

We have examined various approaches to community and dense subgraph detection. What they all have in common is they assign nodes into non-overlapping groups. That is, nodes are either in one group or another but cannot belong to multiple groups at once. While this may make sense for a lot of substantive settings, it might not make sense for other ones, where multiple group memberships are normal (e.g., think of high school). 

## Detecting Overlapping Communities

Methodologically, overlapping community detection methods are not as well-developed as classical community detection methods. Here, we review one simple an intuitive approach that combines the idea of clustering nodes by computing a quantity on the links but instead of computing a rank order (like Newman and Girvan's edge betweenness), we compute pairwise *similarities* between links. We then cluster the links using standard hierarchical clustering methods, which because nodes are incident to many links results in a multigroup clustering of the nodes for free. This approach is called **link clustering** [@ahn_etal10].

Let's see how this works. We first load our trusty *Pulp Fiction* data set from the `networkdata` package, which is an **undirected** graph of character scene co-appearances in the film:

```{r}
    library(networkdata)
    library(igraph)
    library(stringr) #using stringr to change names from all caps to title case
    g <- movie_559
    V(g)$name <- str_to_title(V(g)$name)
    V(g)$name[which(V(g)$name == "Esmarelda")] <- "Esmeralda" #fixing misspelled name
    E(g)$weight <- 1 #setting edge weights to 1.0 (relevant for betweenness calculation)
    g <- subgraph(g, which(degree(g) > 1))
```

The key idea behind link clustering is that *similar* links should be assigned to the same clusters. How do we compute the similarity between links? 

## Measuring Edge Similarity

According to @ahn_etal10 two links $e_{ik}$ and $e_{jk}$ are similar if they share a node $v_k$ *and* the other two nodes incident to each link ($v_i, v_j$) are themselves similar. To measure the similarity between this pair of nodes, we can use any of the off-the-shelf vertex similarity measures that we have seen in action (see lecture notes on [local similarity metrics](sim-loc.qmd)), like Jaccard, cosine, or Dice. @ahn_etal10 use the Jaccard, so that is what we will use.

The first step is thus to build a link by link similarity matrix based on this idea. The following function loops through each *pair of links* in the graph and computes the similarity between two links featuring a common node $v_k$  based on the Jaccard vertex similarity of the two other nodes:


```{r}
    edge.sim <- function(h) {
        E <- ecount(h) #number of edges
        V <- vcount(h) #number of nodes
        if(is.null(V(h)$name) == TRUE) {V(h)$name <- as.character(seq_len(V))} #checking if graph has node names
        edgemat <- ends(h, E(h)) #edge list
        E.sim <- diag(1, E, E) #initializing link similarity matrix to identity matrix
        nlist <- lapply(V(h)$name, function(x) {c(as.vector(names(neighbors(h, x))), x)}) #neighbors list for each node
        names(nlist) <- V(h)$name #naming neighbors list using graph node names
        for (j in 2:E) {
            for (i in 1:(j-1)) { #looping through upper triangle only
                ei <- edgemat[i, ] #nodes incident to first edge
                ej <- edgemat[j, ] #nodes incident to second edge
                if (length(intersect(ei, ej)) != 0) { #do only if pair of edges share a node
                    v <- setdiff(union(ei, ej), intersect(ei, ej))#extracting nodes attached to shared node
                    n1 <- nlist[[v[1]]] #neighbors of first node
                    n2 <- nlist[[v[2]]] #neighbors of second node
                    E.sim[j, i] <- length(intersect(n1, n2)) / length(union(n1, n2)) #Jaccard similarity
                    E.sim[i, j] <- E.sim[j, i] #symmetric score
                    }
                }
            }
        return(E.sim)
        }
```

The function takes the graph as input and returns an inter-link similarity matrix of dimensions $E \times E$ where $E$ is the number of edges in the graph. It works like this:

- In line 5, we extract the end nodes incident to each edge in the graph using the `igraph` function `ends` (which creates an $E \times 2$ matrix the nodes at end of the edges in one column and the nodes at the other end of the edges in the other column). 
- In line 6, we initialize the $E \times E$ matrix of edge similarities to the identity matrix (all edges maximally similar to themselves and have initial similarity zero to other edges). 
- Lines 7 and 8 create a `list` object called `nlist` using the `lapply` shortcut, containing the neighbors of each node in the graph as elements (including the ego node itself, the so-called **inclusive neighborhood**).
- Then lines 9-21 enter a double `for` loop that cycles through each pair of edges, determines whether they have a node in common (line 13) and calculates the Jaccard similarity between the neighborhoods of the other two nodes using the info stored in the `nlist` object (lines 15-16), and fills that info into the edge similarity matrix (lines 17-18).

Here's the function in action in the *Pulp Fiction* network:

```{r}
    E.sim <- edge.sim(g)
    round(E.sim[1:10, 1:10], 3)
```

We can then transform the link similarities into distances, and cluster them:

```{r}
    D <- 1 - E.sim
    D <- as.dist(D)
    hc.res <- hclust(D, method = "ward.D2")
```

The resulting dendrogram looks like this:

```{r}
    par(cex = 0.5)
    plot(hc.res)
```

The leaves of the dendrogram (bottom-most objects) represent each link in the graph ($E = 102$ in this case), and the clusters are "link communities" [@ahn_etal10].

## Clustering Nodes

As we noted, because it is the links that got clustered, the nodes incident to each link can go to more than one cluster (because nodes with degree $k>1$ will be incident to multiple links). 

We can then use the dendrogram information to return a list of node assignments to multiple communities. Let's see the results with ten overlapping communities:

```{r}
    k <- 10
    lc <- cutree(hc.res, k)
    ld <- data.frame(as_edgelist(g), lc)
    node.clus <- list() 
    for (k in 1:max(lc)) {
        sub.dat <- dplyr::filter(ld, lc == k)
        node.clus[[k]] <- unique(c(sub.dat[, 1], sub.dat[, 2]))
        }
    node.clus
```

Because there are nodes that belong to multiple communities, the resulting actor by community ties form a **two-mode network** (see [here](basic-tm.qmd)). 

We can re-construct the **bi-adjacency matrix** corresponding to this two-mode network from the list of community memberships for each node like this:

```{r}
    B <- matrix(0, vcount(g), k)
    rownames(B) <- V(g)$name
    for (i in 1:k) {
        B[node.clus[[i]], i] <- 1
        }
    B
```

And the matrix of inter-community ties based on shared characters can be obtained via the usual @breiger74 projection:

```{r}
    M <- t(B) %*% B
    M
```

And we can visualize the nodes connected to multiple communities as follows: 

```{r}
#| label: fig-overlap
#| fig-cap: "Undirected Pulp Fiction Network Showing the Memberships of Each Node in Overlapping Communities (Nodes who Belong to Non-overlapping Communities Omitted)."
#| fig-height: 13
#| fig-width: 13
#| echo: false
    W <- B/rowSums(B)
    W <- split(W, seq(nrow(W)))
    ccol <- lapply(W, function(x) {which(x != 0)})
    W <- lapply(W, function(x) {x[which(x != 0)]})
    library(RColorBrewer)   
    getPalette = colorRampPalette(categorical_pal(8))
    pal = getPalette(ncol(B))
    V(g)$shape <- "pie"
    V(g)$pie <- W
    c.list <- lapply(ccol, function(x) {pal[x]})
    V(g)$pie.color <- c.list
    set.seed(123)
    plot(g, 
        layout = layout_with_fr,       
        vertex.size = 8,             
        vertex.label.color = "black", 
        vertex.label.font = 2,      
        vertex.label.cex = 1.5,   
        edge.width = 2,                
        edge.color = "lightblue",
        vertex.label.dist = 1.2,
        vertex.label.degree = 180
        )
    legend("bottomright", 
       legend = c(paste("Commnunity", 1:k, sep = "_")), 
       fill = pal, 
       bty = "n",  # No box around legend
       cex = 1.5)  # Text size
```






