---
title: "Using Edge Clustering to Assign Nodes to Overlapping Communities "
execute: 
  eval: true
  echo: true
  output: true
  warning: false
  message: false
format: 
   html:
      code-line-numbers: true
---

We have examined various approaches to community and dense subgraph detection. What they all have in common is they assign nodes into non-overlapping groups. That is, nodes are either in one group or another but cannot belong to multiple groups at once. While this may make sense for a lot of substantive settings, it might not make sense for other ones, where multiple group memberships are normal (e.g., think of high school). 

## Detecting Overlapping Communities

Methodologically, overlapping community detection methods are not as well-developed as classical community detection methods. Here, we review one simple an intuitive approach that combines the idea of clustering nodes by computing a quantity on the links but instead of computing a rank order (like Newman and Girvan's edge betweenness), we compute pairwise *similarities* between links. We then cluster the links using standard hierarchical clustering methods, which because nodes are incident to many links results in a multigroup clustering of the nodes for free. This approach is called **link clustering** [@ahn_etal10].

Let's see how this works. First, let's load up the friendship nomination network from @lazega01, constrained to be undirected:

```{r}
    library(networkdata)
    library(igraph)
    g <- law_friends
    g <- subgraph(g, degree(g) >= 2)
    g <- as_undirected(g, mode = "collapse")
    V(g)$name <- as.character(1:vcount(g))
```

The key idea behind link clustering is that *similar* links should be assigned to the same clusters. How do we compute the similarity between links? 

## Measuring Edge Similarity

According to @ahn_etal10 two links $e_{ik}$ and $e_{jk}$ are similar if they share a node $v_k$ *and* the other two nodes incident to each link ($v_i, v_j$) are themselves similar. To measure the similarity between this pair of nodes, we can use any of the off-the-shelf vertex similarity measures that we have seen in action (see lecture notes on [local similarity metrics](sim-loc.qmd)), like Jaccard, cosine, or Dice.

The first step is thus to build a link by link similarity matrix based on this idea. The following function loops through each *pair of links* in the graph and computes the similarity between two links featuring a common node $v_k$  based on the Jaccard vertex similarity of the two other nodes:

```{r}
    edge.sim <- function(x) {
      el <- as_edgelist(x) #edgelist representation
      A <- as.matrix(as_adjacency_matrix(x))
      S <- A %*% A #shared neighbors
      d <- degree(x)
      E <- nrow(el)
      E.sim <- matrix(0, E, E) #initializing link similarity matrix
      for (e1 in 1:E) {
        for (e2 in 1:E) {
            if (e1 < e2 & length(intersect(el[e1, ], el[e2])) != 0) {
                v <- setdiff(union(el[e1,], el[e2,]), intersect(el[e1,], el[e2,]))
                p <- S[v[1], v[2]]
                q <- d[v[1]] - p
                r <- d[v[2]] - p
                E.sim[e1, e2] <- p / sqrt((p + q) * (p + r))
                E.sim[e2, e1] <- E.sim[e1, e2]          
                }
            }
          }
        return(E.sim)
        }
```

The function takes the graph as input and returns an inter-link similarity matrix of dimensions $E \times E$ where $E$ is the number of edges in the graph:

```{r}
    E.sim <- edge.sim(g)
    round(E.sim[1:10, 1:10], 3)
```

We can then transform the link similarities into distances, and cluster them:

```{r}
    D <- 1 - E.sim
    D <- as.dist(D)
    hc.res <- hclust(D, method = "ward.D2")
```

The resulting dendrogram looks like this:

```{r}
    par(cex = 0.5)
    plot(hc.res)
```

The leaves of the dendrogram (bottom-most objects) represent each link in the graph ($E = 102$ in this case), and the clusters are "link communities" [@ahn_etal10].

## Clustering Nodes

As we noted, because it is the links that got clustered, the nodes incident to each link can go to more than one cluster (because nodes with degree $k>1$ will be incident to multiple links). 

We can then use the dendrogram information to return a list of node assignments to multiple communities. Let's see the results with twelve overlapping communities:

```{r}
    k <- 30
    lc <- cutree(hc.res, k)
    ld <- data.frame(as_edgelist(g), lc)
    node.clus <- list() 
    for (k in 1:max(lc)) {
        sub.dat <- dplyr::filter(ld, lc == k)
        node.clus[[k]] <- unique(c(sub.dat[, 1], sub.dat[, 2]))
        }
    node.clus
```

Because there are nodes that belong to multiple communities, the resulting actor by community ties form a **two-mode network**. 

We can re-construct the bi-adjacency matrix corresponding to this network from the list of community memberships for each node like this:

```{r}
    B <- matrix(0, vcount(g), k)
    rownames(B) <- V(g)$name
    for (i in 1:k) {
        B[node.clus[[i]], i] <- 1
        }
    B
```

And the matrix of inter-community ties based on shared characters can be obtained via the usual @breiger74 projection:

```{r}
    M <- t(B) %*% B
    M
```

And we can visualize the nodes connected to multiple communities as follows: 

```{r}
#| label: fig-overlap
#| fig-cap: "Undirected Pulp Fiction Network Showing the Memberships of Each Node in Overlapping Communities (Nodes who Belong to Non-overlapping Communities Omitted)."
#| fig-height: 12
#| fig-width: 12
#| echo: false
    one.comm <- which(rowSums(B) != 0)
    M <- B[one.comm, ]
    gc <- subgraph(g, one.comm)
    W <- M/rowSums(M)
    W <- split(W, seq(nrow(W)))
    ccol <- lapply(W, function(x) {which(x != 0)})
    W <- lapply(W, function(x) {x[which(x != 0)]})
    names(W) <- V(gc)$name
    library(RColorBrewer)   
    getPalette = colorRampPalette(categorical_pal(8))
    pal = getPalette(ncol(B))
    V(gc)$shape <- "pie"
    V(gc)$pie <- W
    c.list <- lapply(ccol, function(x) {pal[x]})
    V(gc)$pie.color <- c.list
    set.seed(123456)
    plot(gc, 
        layout = layout_with_kk,       
        vertex.size = 12,             
        vertex.label.color = "white", 
        vertex.label.font = 2,      
        vertex.label.cex = 1.5,   
        edge.width = 2,                
        edge.color = "lightblue",
        )
    legend("bottomleft", 
       legend = c(paste("C", 1:k, sep = "_")), 
       fill = pal, 
       bty = "n",  # No box around legend
       cex = 1.2)  # Text size
```






