---
title: "SimRank"
execute: 
  eval: true
  echo: true
  output: true
  warning: false
  message: false
format: 
   html:
      code-line-numbers: true
---


So far, we have defined distances and similarities mainly as a function of the number of shared neighbors between two nodes. Structural equivalence is an idealized version of this, obtaining whenever people share exactly the same set of neighbors.

Yet, similarities based on local neighborhood information only have been criticized (e.g., by @borgatti_everett92) for not quite capturing the sociological intuition behind the idea of a *role* which is usually what they are deployed for. 

That is, two doctors don't occupy the same role because they treat the *same* set of patients (in fact, this would be weird); instead, the set of doctors occupy the same role to the extent they treat a set of actors in the same (or similar) role: Namely, patients, regardless of whether those actors are literally the same or not.  Note that all the similarity notions studied in the [local similarity lecture](locsim.qmd) have structural equivalence as the limiting case of maximum similarity. So they are still based on the idea that *nodes are similar to the extent they connect to the same others*. 

This worry led social network analysts to try to generalize the concept of structural equivalence using less stringent algebraic definition, resulting in something of a proliferation of different equivalences such as **automorphic** or **regular equivalence** [@borgatti_everett92]. Unfortunately, none of this work led (unlike the work on structural equivalence) to useful or actionable tools for data analysis, with some even concluding that some definitions of equivalence (like regular equivalence) are unlikely to be found in any interesting substantive setting. 

A better approach here is to use the more relaxed notion of similarity to come up with a more general definition that can capture our role-based intuitions. One strategy is to define similarity in a more "generalized" way as follows: *nodes are similar to the extent they connect to similar others*, with the restriction that we can only use endogenous (structural connectivity) information---like with structural equivalence or common-neighbor approaches---to define everyone's similarity (no exogenous attribute stuff).

## Jeh and Widom's SimRank

As @jeh_widom02 note, this approach does lead to a coherent generalization of the idea of similarity for nodes in graphs because we can just define similarity *recursively* and iterate through the graph until the similarity scores stop changing.^[Note the similarity (heh, heh) between this idea and the various status ranking algorithms for computing [prestige scores](tm-prestige.qmd).] 

More specifically, they propose to measure the similarity between two nodes $i$ and $j$ at each iteration-step $t$ ($s_{ij}(t)$) using the formula:


$$
   s_{ij}(t) = \frac{\alpha}{d_i d_j} \sum_{k \in N(i)} \sum_{l \in N(j)} s_{kl}(t-1)
$$ {#eq-simrank}

So the similarity between two nodes at step $t$ is just the sum of the pairwise similarities between each of their neighbors (computed in the previous step, $t-1$), weighted by the ratio of a free parameter $\alpha$ (a number between zero and one) to the product of their in-degrees (to take a weighted average). 

This measure nicely captures the idea that nodes are similar to the extent they are both connected to people who are themselves similar. Note that it doesn't matter whether these in-neighbors are shared between the two nodes (the summation occurs over each pair of nodes formed by crossing the set $p$ + $q$ against $p$ + $r$ as defined in the [local similarity](locsim.qmd) lecture), whether they are themselves neighbors, which deals with the doctor/patient problem we referred to earlier. 

A function that implements this idea looks like:

```{r}
   SimRank <- function(w, alpha = 0.8) {
      d <- colSums(w) #indegree vector
      n <- nrow(w) #number of nodes in the graph
      z <- diag(1, n, n) #diagonal matrix (nodes are maximally similar to themselves)
      if(is.null(rownames(w)) == TRUE) {rownames(w) <- 1:nrow(w)} #checking if matrix has row names
      if(is.null(colnames(w)) == TRUE) {colnames(w) <- 1:nrow(w)} #checking if matrix has column names
      rownames(z) <- rownames(w) #naming similarity matrix rows
      colnames(z) <- colnames(w) #naming similarity matrix columns
      nei <- apply(w, 2, function(x){which(x == 1)}) #in-neighbors list
      delta <- 1
      while(delta > 1e-10) { #run while delta is bigger than a tiny number
         z.o <- z #old similarity matrix at t-1
         for(i in 1:n) {
            for(j in 1:n) {
               if (i != j) { 
                  sigma <- 0 #initialize ij sum for iteration t
                  for (k in nei[[i]]) { #loop through i's neighbors
                     for (l in nei[[j]]) { #loop through j's neighbors
                        sigma <- sigma + z[k, l] #i's similarity to j is equal to sum of neighbors' similarity
                        } #end l for loop
                     } #end k for loop
                     z[i,j] <- alpha/(d[i] * d[j]) * sigma #updating ij similarity score at t
                  } #end if
               } #end j for loop
            } #end i for loop
         delta <- abs(sum(abs(z)) - sum(abs(z.o))) #difference between similarity matrices at t and t-1
         } #end while loop
      return(z) #return similarity matrix after convergence
      } #end function
```

This function takes an adjacency matrix as input (`w`) and returns a generalized similarity matrix between nodes (`z`) as output.

Let us see how it works using the toy example that @jeh_widom02 [p. 540, Figure 2] used in their original paper, shown in @fig-jw, which is supposed to be a set of web pages in an university. There is a link from the main University page to that of the two professors, and a link from the professor to the student pages. In addition, "Student B" has a link on their webpage to the corresponding professor, and "Student A" has a link on their webpage to the main university page. 

```{r}
#| label: fig-jw
#| fig-cap: "Jeh and Widom's Toy Example"
#| fig-height: 12
#| fig-width: 12
#| echo: false
   library(igraph)
   g <- make_empty_graph(5, directed = TRUE)
   V(g)$name <- c("Univ", "ProfA", "ProfB", "StudentA", "StudentB")
   g <- g + edge("Univ","ProfA")
   g <- g + edge("Univ","ProfB")
   g <- g + edge("ProfA","StudentA")
   g <- g + edge("ProfB", "StudentB")
   g <- g + edge("StudentB", "ProfB")
   g <- g + edge("StudentA", "Univ")
   plot(g, 
      vertex.size=14, vertex.frame.color="lightgray", 
      vertex.label = V(g)$name, edge.arrow.size = 2, 
      layout = layout.auto, vertex.label.cex = 2, 
       vertex.label.dist=2)
```

We can now test the `SimRank` function on the toy example:

```{r}
   A <- as.matrix(as_adjacency_matrix(g))
   S <- SimRank(A)
   round(S, 3)
```

Which reproduces the pairwise similarity numbers reported by @jeh_widom02 [p. 540, Figure 2] in their original paper.


We can transform the similarity matrix into a distance matrix and then perform a hierarchical clustering to obtain the roles:

```{r}
   D <- as.dist(1-S)
   h.res <- hclust(D, method = "ward.D") #hierarchical clustering
   plot(h.res)
```

Which, as we can see, recovers the Professor, Student, and University Roles in the network. 

Let's try it out on more realistic data using the *Pulp Fiction* graph:

```{r}
   library(networkdata)
   library(stringr) #using stringr to change names from all caps to title case
   g <- movie_559
   V(g)$name <- str_to_title(V(g)$name)
   V(g)$name[which(V(g)$name == "Esmarelda")] <- "Esmeralda" #fixing misspelled name
   g <- delete_vertices(g, degree(g) < 2) #deleting low degree vertices
   A <- as.matrix(as_adjacency_matrix(g))
   S <- SimRank(A)
   round(S[1:5, 1:5], 2)
```

We can transform the generalized similarities to distances and plot:

```{r}
#| fig-height: 10
#| fig-width: 10
   D <- as.dist(1- S)
   h.res <- hclust(D, method = "ward.D2") #hierarchical clustering
   plot(h.res)
```

Note that the lowest levels of the dendrogram hierarchy some of the pairings make a lot of sense: The Gimp is most similar to Zed, Pumpkin is most similar to Honey Bunny (and other characters in the diner scene), Jimmie is most similar ot the Wolf, and Jules, Marsellus, Butch, Mia, and Vincent (the main characters) are most similar to one another. 

The plot suggests a division into five major node clusters. We can obtain that as follows using the `cutree` function:

```{r}
   blocks  <- cutree(h.res, k = 5)
   blocks
```

We can see that the people in the diner scene all end up in the same cluster (4) which makes sense, as do Maynard, Zed, and the Gimp (5) which also makes sense. The main characters also end up in the same generalized similarity cluster (1), suggesting they play a similar role in the story. 

## Appendix

A more general (and shorter) version of the function to compute `SimRank` looks like this [partly based on @fouss_etal16, pp. 84, algorithm 2.4]:

```{r}
   SimRank2 <- function(w, alpha = 0.85) {
      n <- nrow(w) #number of nodes
      d <- as.numeric(colSums(w) > 0) #degree vector
      e <- matrix(1, n, 1) #all ones column vector
      q <- diag(as.vector(t(e) %*% w), n, n) #q matrix
      diag(q) <- 1/diag(q) #setting q diagonals to inverse
      q <- w %*% q
      q[is.nan(q)] <- 0
      delta <- 1
      k <- diag(1, n, n)
      while(delta > 1e-10) {
         k.o <- k
         k.p <- alpha * t(q) %*% k %*% q
         k <- k.p - diag(diag(k.p), n, n) + diag(d, n, n)
         delta <- abs(sum(abs(k)) - sum(abs(k.o)))
         }
      rownames(k) <- rownames(w)
      colnames(k) <- colnames(w)
      return(k)
   }
```