---
title: "Status and Prestige"
execute: 
  eval: true
  echo: true
  output: true
  warning: false
  message: false
format: 
   html:
      code-line-numbers: true
---

In the last handout, we saw how to compute the most popular centrality measures. Freeman's "big three" have strong graph-theoretic foundation and do a good job of formalizing and quantifying the idea that a node is central if it is "well-placed" in the network, where being well-placed resolves into either being able to *reach* others (directly as with degree or indirectly as with closeness) or being able to *intermediate* between others (as with betweenness).

## Networks as Prisms

There is, however, another strong and well-motivated intuition as to what it means to be "well-placed" in a network. Here the ties in the network are seen less as "pipes" that transmit stuff and more like "prisms" that *reflect* on you [@podolny01]. Another way of thinking about this second version of well-placedness is that what is transmitted through the network is the *network itself*, or more accurately, the *importance*, *status*, and *prestige* of the people you are connected to.

Under this interpretation, actors get status and prestige in the network from being connected to prestigious and high status others. Those others, in turn, get their status from being connected to high status others, and so *ad infinitum*. 

One way of quantifying this idea goes like this. If $\mathbf{x}$ is a vector containing the desired status scores, then the status of each actor should be equal to:

$$
   x_i = \sum_{j} a_{ij}x_j
$$ {#eq-status-sum}

Where $a_{ij} = 1$ if $i$ is adjacent to $j$ in the network. Note that this formula just sums up the status scores of all the others each actor is connected to.

In matrix notation, if $\mathbf{x}$ is a column vector of status scores then:

$$
   \mathbf{x} = A\mathbf{x}
$$

Because $\mathbf{x}$ is an $n \times n$ matrix and $\mathbf{x}$ is $n \times 1$ column vector, the multiplication $A\mathbf{x}$ will return another column vector of dimensions $n \times 1$, in this case $\mathbf{x}$ itself!

Note the problem that this formulation poses: $\mathbf{x}$ appears on both sides of the equation, which means that in order to know the status of any one node we would need to know the status of the others, but calculating the status of the others depends on knowing the status of the focal node, and so on. There's a chicken and the egg problem here.

Now, there is an obvious (to the math majors) *mathematical solution* to this problem, because there's a class of solvable (under some mild conditions imposed on the matrix $\mathbf{A}$) linear algebra problems that take the form:

$$
   \lambda\mathbf{x} = A\mathbf{x}
$$

Where $\lambda$ is just a plain old number (a scalar). Once again conditional of the aforementioned mild conditions being met, we can search for a value $\lambda$ while also filling up the $\mathbf{x}$ with another set of values until we find some set of values for both $\lambda$ and the entries in $\mathbf{x}$ that make the above equality true. 

When we do that successfully, we say that the value of $\lambda$ we hit upon is an **eigenvalue** of the matrix $\mathbf{A}$ and the values of the vector $\mathbf{x}$ we came up with are an **eigenvector** of the same matrix (technically in the above equation a right eigenvector). 

Eigenvalues and eigenvectors, like Don Quixote and Sancho Panza, come in pairs, because you need a unique combination of both to solve the equation. Typically, a given matrix (like an adjacency matrix) will have multiple $\lambda/\mathbf{x}$ pairs that will solve the equation. Together the whole set $\lambda/\mathbf{x}$ pairs that make the equation true are the **eigenvalues** and **eigenvectors** of the matrix. 

Note that all of this obscure talk about eigenvalues and eigenvectors is just matrix math stuff. It has nothing to do with networks and social structure. In contrast, because the big three centrality measures have a direct foundation in graph theory, and graph theory is an isomorphic **model** of social structures (points map to actor and lines map to relations) the "math" we do with graph theory is **directly** meaningful as a model of networks (the counts of the number of edges incident to a node is the count of other actors they someone is directly connected to). 

Eigenvalues and eigenvectors are not a model of social structure in the way graph theory is (their first scientific application was in Physics). They are just a mechanical math fix to a circular equation problem. 

This is why it's a mistake to introduce network measures of status and prestige by jumping directly to the machinery of linear algebra (or worse talk about the idea of **eigenvector centrality** which means nothing to most people). 

A better approach is to see if we can *motivate* the use of measures like the ones above using the simple model of the distribution of status and prestige we started with earlier. We will see that we can, and that doing that leads us back to solutions that are the mathematical equivalent of all the eigenvector stuff. 

## Distributing Centrality to Others

Let's start with the simplest model of how people can get their status from the status of others in a network. It is the simplest because it is based on degree. 

Imagine everyone has the same "quantum" of status to begin with (this can be stored in a vector containing the same number of length equals to number of actors in the network). Then, at each step, people "send" the same amount of status to all their alters in the network. At the end of each step, we compute people's status scores using @eq-status-sum. We stop doing this after the status scores of people stop changing across each iteration.

Let us see a real-life example at work. 

We will use a data set collected by David Krackhardt on the friendships of 21 managers in a high tech company in the West coast. The data are reported as directed ties ($i$ nominates $j$ as a friend) but we will constrain ties to be undirected:

```{r}
   library(networkdata)
   library(igraph)
   g <- as.undirected(ht_friends)
```

This is what the network looks like:

```{r}
#| echo: false
#| fig-cap: "Krackhardt's Manager Data."
#| fig-height: 7
   library(ggraph)
   set.seed(456)
    p <- ggraph(g, layout = 'dh')
    p <- p + geom_edge_fan(color = gray(0.5), edge_width = 0.25)
    p <- p + geom_node_point(aes(x = x, y = y), size = 8, color = "tan2") 
    p <- p + geom_node_text(aes(label = 1:vcount(g)), size = 4, color = "white")
    p <- p + theme_graph() 
    p
```

We extract the adjacency matrix corresponding to this network:

```{r}
   A <- as.matrix(as_adjacency_matrix(g))
```

And here's a simple custom function using a `while` loop that exemplifies the process of status distribution through the network we talked about earlier:

```{r}
   status1 <- function(A) {
      n <- nrow(A) #number of actors
      x <- rep(1, n) #initial status vector set to all ones
      w <- 1 
      k <- 0 #initializing counter
      while (w > 0.0001) {
          o.x <- x #old status scores
          x <- A %*% x #new scores a function of old scores and adjacency matrix
          x <- x/norm(x, type = "E") #normalizing new status scores
          w <- abs(sum(abs(x) - abs(o.x))) #diff. between new and old scores
          k <- k + 1 #incrementing while counter
      }
   return(as.vector(x))
   }
```

Lines 1-5 initialize various quantities, most importantly the initial status vector for each node to just a series of ones:

```{r}
   rep(1, nrow(A))
```

Then lines 6-12 implement a little loop of how status is distributed through the network, with the most important piece of code being line 8 where the *current* status scores for each node are just the sum of the status scores of its neighbors computed one iteration earlier. The program stops when the difference between the old and the new scores is negligible as checked in line 10. 

Note the normalization step on line 9, which is necessary to prevent the sum of status scores from getting bigger and bigger indefinitely. The `type = "E"` normalization is called the **euclidean vector norm**, which we divide each value of the status scores by after each update. For a vector of numbers $\mathbf{x}$ the euclidean vector norm $||\mathbf{x}||_2$ is given by: $\sqrt{\sum x^2}$. So we divide each status scores by the square root of the sum of all the squared scores.

And here's the resulting (row) vector of status scores for each node:

```{r}
   round(status1(A), 3)
```      

What if I told you that this vector is the same as that given by the leading (first) eigenvector of the adjacency matrix?

```{r}
   eig <- eigen(A)
   round(eig$vector[, 1], 3) * -1
```

Which is of course what is computed by the `eigen_centrality` function in `igraph`:

```{r}
   round(eigen_centrality(g, scale = FALSE)$vector, 3)
```

So, the eigenvector centralities are just the limit scores produced by the status distribution process implemented in the `status1` function!

Most other measures of status in networks are constructed using similar principles. What changes is the *model* of how status is distributed in the system. That's why scary and non-intuitive stuff about eigenvectors or whatever is misleading.

Other measures are constructed such that they either change the quantum of status that is distributed through the network by making it dependent on some node characteristic (like degree) or differentiate between different routes of distribution in **directed graphs**, by for instance, differentiating status derived from outgoing links from that derived from incoming links. 

Let's see some other examples.

## A Degree-Dependent Model of Status

Note that the model of status distribution implied by the traditional **eigenvector centrality** implies that each node distributes the same amount of status *independently* of the number of connection it has; status just replicates. Thus, a node with a 100 friends has 100 status units to distribute to each of them and a node with a 10 friends has 10 units. 

This is why the eigenvector idea rewards nodes who are connected to popular others more. Even though everyone begins with a single unit of status, well-connected nodes by degree end up having more of it to distribute. 

But what if status dissipated proportionately to the number of connections one had? For instance, if someone has 1/100 friends and they only had so much time or energy, they would only have a fraction of status to distribute to others than a person with 10 friends. 

In that case, the node with a hundred friend would only have 1/100 of status unites to distribute to each of their connections while the node with 10 friends would have 1/10 units. Under this formulation, being connected to *discerning* others, that is people who only connect to a few, is better than being connected to others who connect to everyone else indiscriminately. 

How would we implement this model? First, let's create a variation of the adjacency matrix called the $\mathbf{P}$ matrix:

```{r}
   P <- A/rowSums(A)
```

So this is the original adjacency matrix, with each entry $a_{ij}$ divided by the sum of the corresponding row, which, as you may recall, is equivalent to the **degree** of node $i$. 

Here are the first 10 rows and columns of the new matrix:

```{r}
   round(P[1:10, 1:10], 2)
```

Note that the entries are now numbers between zero and one and the matrix is *asymmetric* that is $p_{ij}$ is not necessarily equal to $p_{ji}$. In fact $p_{ij}$ will only be equal to $p_{ji}$ when $k_i = k_j$ (nodes have the same degree). 

Moreover the rows of $\mathbf{P}$ sum to one:

```{r}
   rowSums(P)
```

Which means that the $\mathbf{P}$ matrix is **row stochastic**. That is the "outdegree" of each node in the matrix is forced to sum to a fixed number (which means that it is a useless quantity). However, the indegree is not:

```{r}
   round(colSums(P), 2)
```

Which means that inequalities in the system will be tied to the indegree of each node in the $\mathbf{P}$ matrix. This matrix has many interpretations, but here it just quantifies the idea that the amount of centrality each node can distribute is proportional to their degree, and that the larger the degree, the less there is to distribute (the smaller each cell $p_{ij}$ will be). Meanwhile, it is clear that nodes that are pointed to by many other nodes who themselves don't point to many others have a large indegree in $\mathbf{P}$. 

Now we can just write function implementing the model of status distribution we used for eigenvector centrality but this time using the $\mathbf{P}$ rather than the $\mathbf{A}$ matrix. Note that because we interested in the status that comes **into** each node we use the *transpose* of $\mathbf{P}$ rather than $\mathbf{P}$. So at each step the status of a node is equivalent to the sum of the status scores of their in-neighbors, with more discerning in-neighbors passing along more status than less discerning ones:

```{r}
   round(status1(t(P)), 3)
```

What if I told you that these numbers are the same as the leading eigenvector of $\mathbf{P}^T$?

```{r}
   eig <- eigen(t(P))
   round(eig$vector[, 1], 3) * -1
```

And, of course, the (normalized) scores produced by this approach are identical to those computed by the `page_rank` function in `igraph`:

```{r}
   s2 <- status1(t(P))
   pr <- page_rank(g, damping = 1, algo = "arpack")$vector
   round(pr/max(pr), 3)
   round(s2/max(s2), 3)
```

So the distributional model of status is the same one implemented in the PageRank algorithm!

PageRank of course was designed to deal with **directed graphs** (like the World Wide Web). So let's load up the version of the Krackhardt's Managers data that contains the **advice network** which is an unambiguously directed relation. 

```{r}
   g <- ht_advice
```

We then compute the $\mathbf{P}$ matrix:

```{r}
   A <- as.matrix(as_adjacency_matrix(g))
   P <- A/rowSums(A)
   round(P, 2)
```

Remember how we said earlier that there are multiple ways of thinking about $\mathbf{P}$? Another way of thinking about the $\mathbf{P}$ matrix is as characterizing the behavior of a **random walker** in the directed graph. At any time point $t$ the walker (a piece of information, a virus, or status itself) sits on a node and the with probability $p_{ij}$ it jumps to one of that node's out-neighbors. The probabilities are stored in the matrix $\mathbf{P}$. 

One issue that arises is that there could be nodes with no out-neighbors (so-called sink nodes) or like node 6 in the above matrix nodes, with just one out-neighbor, in which case the probability is 1.0 that if the random walker is at node 6 it will go to node 21. 

To avoid this issue the original designers of the PageRank algorithm [@brin_page98] added a "fudge" factor: That is, with probability $\alpha$ the random walker should hop from node to node following the directed links in the graph. But once in a while with probability $1-\alpha$ the walker should decide to "teleport" (with uniform probability) to *any* node in the graph.

How do we do that? Well we need to "fix" the $\mathbf{P}$ to allow for such behavior. So instead of $\mathbf{P}$ we estimate our distributive status model on the matrix $\mathbf{G}$ (yes, for Google):

$$
   \mathbf{G} = \alpha \mathbf{P} + (1 - \alpha) \mathbf{E}
$$

Where $\mathbf{E}$ is a matrix of the same dimension as $\mathbf{P}$ but containing $1/n$ in every cell indicating that every node has an equal chance of being teleported to.

So fixing $\alpha = 0.85$ our $\mathbf{G}$ matrix would be:

```{r}
   E <- matrix(1/vcount(g), vcount(g), vcount(g))
   G <- (0.85 * P) + (0.15 * E)
```

And then we just play our status distribution game on the transpose of $\mathbf{G}$:

```{r}
   s3 <- round(status1(t(G)), 3)
   round(s3/max(s3), 3)

```

Which is the same answer you would get from the `igraph` function `page_rank` by setting the "damping" parameter to 0.85:

```{r}
   pr <- page_rank(g, damping = 0.85, algo = "arpack")$vector
   round(pr/max(pr), 3)
```

## Hubs and Authorities

Recall from previous discussions that everything doubles (sometimes quadruples like degree correlations) in directed graphs. The same goes for status as reflected in a distributive model through the network. 

Consider two ways of showing your status in a system governed by directed relations (like advice). You can be highly sought after by others (be an "authority"), or you can be discerning in who you seek advice from, preferably seeking out people who are also sought after (e.g., be a "hub" pointing to high-quality others). 

These two forms of status are mutually defining. The top authorities are those who are sought after by the top hubs, and the top hubs are the ones who seek the top authorities! 

So this leads to a doubling of @eq-status-sum:

$$  
   x^h_i = \sum_j a_{ij} x^a_j
$$

$$
   x^a_i = \sum_i a_{ij} x^h_i
$$

Which says that the hub score $x^h$ of a node is the sum of the authority scores $x^a$ of the nodes they point to, and the authority score of a node is the sum of the hub scores of the nodes that point to it. 

So we need to make our status game a bit more complicated (but not too much) to account for this duality:

```{r}
   status2 <- function(A) {
     n <- nrow(A)
     a <- rep(1, n)  #initializing authority scores
     diff.a.h <- 1 #initializing diff.
     while (diff.a.h >= 0.0001) {
         o.a <- a #old authority scores
         h <- A %*% o.a #new hub scores a function of previous authority scores
         h <- h/norm(h, type = "E")
         a <- t(A) %*% h #new authority scores a function of current hub scores
         a <- a/norm(a, type = "E")
         diff.a.h <- abs(sum(abs(o.a) - abs(a))) #diff. between old and new scores
         }
   return(list(h = as.vector(h), a = as.vector(a)))
   }
```

Everything is like our previous `status1` function except now we are keeping track of two mutually defining scores `a` and `h`. 

The results for the Krackhardt advice network are:

```{r}
   round(status2(A)$a/max(status2(A)$a), 3)
   round(status2(A)$h/max(status2(A)$h), 3)
```

Which are equivalent to using the `igraph` functions `authority_score` and `hub_score`:

```{r}
   round(authority_score(g, scale = TRUE)$vector, 3)
   round(hub_score(g, scale = TRUE)$vector, 3)
```



