---
title: "Statistical Models of Networks III"
execute: 
  eval: true
  echo: true
  output: true
  warning: false
  message: false
format: 
   html:
      code-line-numbers: true
---

In the previous handout we saw how to fit exponential random graph models (ergms) for simple undirected networks. As noted, these models are useful for *multivariate null hypothesis testing* using network data, answering questions of the form: Is this network statistic larger or smaller than I would expect by chance, after conditioning (e.g., "controling for") these *other* network statistics? 

The ergm modeling framework is pretty flexible, allowing you to test for the statistical significance of *any* network statistic you can imagine while holding constant a bunch of other ones.

Here we will see how to extend the ergm framework to other types of directed, two-mode, and temporal networks.

## ERGMs in Directed Networks

The extension to the directed case is the most straightforward.

The key network statistic that we need to think about in the directed case is **reciprocity** or **mutuality**, namely, the fact that in the directed case, it is possible that I send a tie to you but you don't send it back. In this advice network case this would happen if I seek advice from you, but you don't seek advice from me.

In an ergm, we use the `mutual` term to condition on the number of dyads featuring arcs going on both directions (e.g., from $i$ to $j$ *and* from $j$ to $i$).

Let's see an example using Krackhardt's High-Tech Managers:
   
```{r}
   library(networkdata)
   library(igraph)
   library(ergm)
   library(intergraph)
   g1 <- ht_friends
   n1 <- asNetwork(g1)
   g2 <- ht_advice
   n2 <- asNetwork(g2)
   m1 <- ergm(n1 ~ edges + mutual)
   summary(m1)
   m2 <- ergm(n2 ~ edges + mutual)
   summary(m2)
```

Here we see strong evidence for mutuality effects in *friendship* nominations ($\theta_m = 1.33, p < 0.01$), consistent with the social meaning of the relation, but not in advice relations ($\theta_m = 0.13, p = 0.65$), suggesting that these are more hierarchical).

We can also incorporate homophily based non attributes into our analysis of reciprocity. That is, we can test hypotheses of the form: Is mutuality weaker or stronger for dyads that match on a certain characteristic?

For instance, we may suspect that friendship ties among high-tech managers are influenced by formal organizational position, so that there is more mutuality within levels (e.g., mid-manager to mid-manager) than across levels (mid to lower-level). 

Recall than in this network there are three managerial level but only one top manager, so we should compare the middle level to the rest

We can test this hypothesis as using the `same` argument in the `mutual` ergm term:

```{r}
   set.seed(123456)
   m3 <- ergm(n1 ~ edges 
              + mutual(same = "level", diff = TRUE, levels = 2)
            )
   summary(m3)
```

And indeed, we find that compared to lower level managers, tendencies toward mutuality in friendship choices are stronger among mid-level managers. 

We can also test hypotheses of the form: Are nodes with certain characteristics more likely to be involved in mutual dyads *regardless of the characteristics of the node at the other end of the tie*? That is, this tests for differential tendencies toward mutuality based on attribute.

We can do this using the `by` argument in the `mutual` term. This time using department as the relevant attribute:

```{r}
   set.seed(123456)
   m4 <- ergm(n1 ~ edges 
              + mutual(by = "dept", levels = 2:4)
            )
   summary(m4)
```

Which shows that nodes in department 2 tend to have more mutual ties compared to other managers.

In `m3` we found that middle level managers are more likely to form mutual friendship ties to other middle-level managers. We may be interested to see if this effect is net of activity differences based on this factor:


```{r}
   set.seed(123456)
   m5 <- ergm(n1 ~ edges 
              + nodefactor("level", levels = 2)
              + mutual(same = "level", diff = TRUE, levels = 2)
            )
   summary(m5)
```

And, indeed, we find that after adjusting for the fact that middle-managers make less overall friendship nominations, tendencies toward mutuality among same-level ties become stronger. 

Also like before, we may be interested in whether our result remains after we condition on core graph level characteristics like the degree distribution using geometrically weighted (`gw`) terms. 

In the directed case, however, we have to worry about both the *in* and *outdegree* distributions, which means two distinct `gw` terms, namely, `gwidegree` and `gwodegree`. It is usually good to introduce one at a time to see if we have to worry about it::

```{r}
   set.seed(123456)
   m6 <- ergm(n1 ~ edges 
              + gwidegree(decay = 0.25, fixed = TRUE)
              + nodefactor("level", levels = 2)
              + mutual(same = "level", diff = TRUE, levels = 2)
            )
   summary(m6)
```

Which shows that there does not seem to be any significant heterogeneity based on indegree in the friendship network. 

Let's now look at outdegree:

```{r}
   set.seed(123456)
   m7 <- ergm(n1 ~ edges 
              + gwodegree(decay = 0.25, fixed = TRUE)
              + nodefactor("level", levels = 2)
              + mutual(same = "level", diff = TRUE, levels = 2)
            )
   summary(m7)
```

The outdegree distribution is indeed heterogeneous, although that does not change our conclusions regarding tendencies toward mutuality among mid-level managers. 

We can then also check to see whether our results remain the same when we condition on graph-level tendencies towards transitivity. 

Here we have different flavors to choose from because in a directed graph, a dyad can be involved in multiple types of relations with a third party (ingoing, outgoing, reciprocal). 

In this specification, we condition on whether a dyad is connected to a third party via reciprocal ties ("Reciprocal Two-Path" or "RTP"):

```{r}
   set.seed(123456)
   m8 <- ergm(n1 ~ edges 
              + gwdsp(decay = 0.25, fixed = TRUE, type = "RTP")
              + nodefactor("level", levels = 2)
              + mutual(same = "level", diff = TRUE, levels = 2)
            )
   summary(m8)
```

And we see that while indeed, having reciprocated ties to a third party increases the chances of a dyad being connected, conditioning on this fact, does not change our conclusions about heightened tendencies toward mutuality among homophilous mid-manager dyads.




