---
title: "The Duality of Persons and Groups"
execute: 
  eval: true
  echo: true
  output: true
  warning: false
  message: false
format: 
   html:
      code-line-numbers: true
---

Recall that in the [one-mode case](prestige.qmd), multiplying the adjacency matrix times its transpose yields the **common neighbors matrix** $\mathbf{M}$:

$$
\mathbf{M} = \mathbf{A}\mathbf{A}^T
$$

As famously noted by @breiger74, doing the same for the affiliation matrix of a two-mode network also returns the common-neighbors matrix, but because objects in one mode can only connect to objects in another mode, this also reveals the **duality of persons and groups**: The connections between people are made up of the groups they share, and the connections between groups are revealed by the groups they share.

Thus, computing the common neighbors matrix for both persons and groups (also called the **projection** of the two-mode network into each of its modes) produces a one-mode similarity matrix between people and groups, where the similarities are defined by the number of objects in the other mode that they share. 

For the people the relevant projection is:

$$
\mathbf{P} = \mathbf{A}\mathbf{A}^T
$$ {#eq-p}

And for the groups:

$$
\mathbf{G} = \mathbf{A}^T\mathbf{A}
$$ {#eq-g}

Let's see how this works out with real data by loading our usual friend the *Southern Women* data:

```{r}
    library(igraph)
    library(networkdata)
    g <- southern_women
    library(stringr) #using stringr to change names from all caps to title case
    V(g)$name <- str_to_title(V(g)$name)
```

In this case, equations @eq-g and @eq-p yield:

```{r}
   A <- as_biadjacency_matrix(g)
   P <- A %*% t(A)
   P
   G <- t(A) %*% A
   G
```

The off-diagonal entries of these square person by person (group by group) matrices is the number of groups (people) shared by each person (group) and the diagonals are the number of memberships of each person (the size of each group/event).

In `igraph` the  relevant function is called `bipartite_projection`. It takes a bipartite graph as an input and returns a list containing `igraph` graph objects of both projections by default:

```{r}
   Proj <- bipartite_projection(g)
   G.p <- Proj[[1]]
   G.g <- Proj[[2]]
```

In the graph objects produced by the `bipartite_projection` function, the actual shared memberships and shared members are stored as an *attribute* of each edge called `weight` used in the plotting code for @fig-projection to set the `edge.width`:

```{r}
   edge_attr(G.p)
   edge_attr(G.g)
```

So to get the weighted projection matrix, we need to type:

```{r}
   as.matrix(as_adjacency_matrix(G.p, attr = "weight"))
```

```{r}
   as.matrix(as_adjacency_matrix(G.g, attr = "weight"))
```
We can also use the weights to draw a weighted graph network plot of people and group projections. All we have to do is set the `edge.with` argument to the value of the edge weight attribute in the corresponding graph:

```{r}
#| fig-height: 12
#| fig-width: 12
#| fig-cap: "One mode projections."
#| label: fig-projection
#| fig-subcap: 
#|   - "Persons"
#|   - "Groups"
#| layout-ncol: 1

   set.seed(123)
   plot(G.p, 
     vertex.size=8, vertex.frame.color="lightgray", 
     vertex.label.dist=2, edge.curved=0.2, 
     vertex.label.cex = 1.5, edge.color = "lightgray",
     edge.width = E(G.p)$weight)

   set.seed(123)
   plot(G.g, 
     vertex.size=8, vertex.frame.color="lightgray", 
     vertex.label.dist=2, edge.curved=0.2, 
     vertex.label.cex = 1.5, edge.color = "lightgray",
     edge.width = E(G.g)$weight)
```

Note that because both `G.p` and `G.g` are **weighted graphs** we can calculate the weighted version of degree for both persons and groups from them (sometimes called the **vertex strength**).

In `igraph` we can do this using the `strength` function, which takes a weighted graph object as input:

```{r}
   strength(G.p)
   strength(G.g)
```

Interestingly, as noted by @faust97 [p. 167], there is a (dual!) mathematical connection between the strength of each vertex in the weighted projection and the centrality of the nodes from the other set they are connected to:

1. For people, the vertex strength is equal to the sum of the sizes of the groups they belong to minus their own degree.

1. For groups, the vertex strength is equal to the sum of the memberships of the people that belong to them, minus their own size.

We can verify this relationship for Evelyn:

```{r}
   sum.size.evelyn <- sum(A["Evelyn", ] * degree(g)[which(V(g)$type == TRUE)]) #sum of the sizes of the groups Evelyn belongs to
   sum.size.evelyn - degree(g)[which(V(g)$name == "Evelyn")]
```

Which is indeed Evelyn's vertex strength. 

Dually, the same relation applies to groups:

```{r}
   sum.mem.6.27 <- sum(A[, "6/27"] * degree(g)[which(V(g)$type == FALSE)]) #sum of the memberships of people in the first group
   sum.mem.6.27 - degree(g)[which(V(g)$name == "6/27")]
```

Which is indeed the vertex strength of the event held on 6/27.

## Visualizing Dual Projections Using the Minimum Spanning Tree

As we just saw projecting the original biadjacency matrix using Breiger's [-@breiger74] approach results in two weighted one mode networks. Because there is an edge between two persons (groups) if they share at least one group (person) the resulting graphs tend to be dense featuring high levels of connectivity between nodes as with @fig-projection. This can make it hard to discern the connectivity structure of the projected networks and detect patterns.

One approach to simplifying the visual projections of two-mode networks is to calculate the resulting weighted graph's **minimum spanning tree**. For any weighted network, the minimum spanning tree is the graph that connects all nodes using the smallest number of edges that form a tree^[Recall that a tree is a connected graph without any cycles.], which happens to be $N-1$ where $N$ is the number of nodes. 

To do this, we can follow **Kruskal's Algorithm**. It goes like this:

- First, we create an edgelist containing each of the edge weights of the one-mode projection graph.
- Then we sort the edgelist by weight in increasing order (smallest weights first).
- Then, we create an undirected empty graph with $N$ nodes.
- Now, we go down the edgelist adding edges to the empty graph one at a time, at each step checking that:
  1. We are not connecting nodes that have already been connected (avoiding **multiedges**).
  1. Additional edges do not create **cycles** (as the resulting graph would be no longer a tree). 

We stop when we have the desired number of edges ($N-1$), meaning all nodes are connected in the tree.

Here's a function that does all of this, taking the weighted projection `igraph` object as input and returning the minimum spanning tree:

```{r}
   make.mst <- function(x) {
      E <- data.frame(as_edgelist(x), w = E(x)$weight) #creating weighted edge list
      E <- E[order(E$w), ] #ordering by edge weight
      Tr <- make_empty_graph(n = vcount(x), directed = FALSE) #creating empty graph
      V(Tr)$name <- V(x)$name
      k <- 1
      n.e <- ecount(Tr)
      n.v <- vcount(Tr) - 1
      while(n.e < n.v) {
         i <- which(V(x)$name == E[k,1])
         j <- which(V(x)$name == E[k,2])
         if (are_adjacent(Tr, i, j) == 0) { #checking nodes are not adjacent
            Tr <- add_edges(Tr, c(i,j)) #add edge
            n.e <- ecount(Tr) #new edge id
            if (is_acyclic(Tr) == 0) { #checking new edge does not add a cycle
               Tr <- delete_edges(Tr, n.e) # delete edge if it adds a cycle
               }
            }
         n.e <- ecount(Tr)
         k <- k + 1
         }
      return(Tr)
   }
```

We can now build the minimum spanning tree graph for each weighted projection:

```{r}
   Tr.p <- make.mst(G.p)
   Tr.g <- make.mst(G.g)
```

And here's a point and line plot the minimum spanning tree for persons and groups in the *Southern Women* data (we use the `layout_as_tree` option in `igraph`):

```{r}
#| fig-height: 12
#| fig-width: 12
#| fig-cap: "One mode projection MST."
#| label: fig-mst
#| fig-subcap: 
#|   - "Persons"
#|   - "Groups"
#| layout-ncol: 1

set.seed(123)
   plot(Tr.p, 
     vertex.size=8, vertex.frame.color="lightgray", 
     vertex.label.dist=2, layout=layout_as_tree,
     vertex.label.cex = 1.5, edge.color = "lightgray",
     edge.width = 3, vertex.color = cluster_leading_eigen(Tr.p)$membership)
set.seed(123)
   plot(Tr.g, 
     vertex.size=8, vertex.frame.color="lightgray", 
     vertex.label.dist=2, layout=layout_as_tree,
     vertex.label.cex = 1.5, edge.color = "lightgray",
     edge.width = 3, vertex.color = cluster_leading_eigen(Tr.g)$membership)
```

## An Asymmetric Approach to Dual Projection

@zhou_etal07 describe an asymmetric approach to projecting a two-mode network into two one-mode weighted networks. The motivation is to move beyond two limitations of the Breiger projection approach: (1) the fact that the standard projection always results in a *symmetric* weighted adjacency matrix, and (2) the fact that nodes in two-mode networks that only connect to single entity in the other node are necessarily excluded from the traditional projection (they are isolates).

To deal with this issue, @zhou_etal07 propose to weigh the one mode projections reciprocally by the degree of the nodes in each mode, similar to the degree-weighting that motivates the CA approach to analyzing two-mode networks we considered [here](tm-ca.qmd). So the idea is that instead of using the usual Breiger projection we considered above, we instead use the *degree-normalized projection*. 

Recall from that [lecture](tm-ca.qmd) that the degree-normalized projections for persons and groups are obtained as follows:

```{r}
   D.p <- diag(1/rowSums(A)) #inverse of degree matrix of persons
   P.pg <- D.p %*% A
   rownames(P.pg) <- rownames(A) 
   D.g <- diag(1/colSums(A)) #inverse of degree matrix of groups
   P.gp <- D.g %*% t(A) #person-degree normalized biadjacency matrix
   rownames(P.gp) <- colnames(A) #group-degree normalized biadjacency matrix
   P.pp <- P.pg %*% P.gp #degree-normalized projection for people
   P.gg <- P.gp %*% P.pg #degree-normalized projection for groups
```

Checking the first rows and columns of the degree-weighted person projection reveals:

```{r}
   round(P.pp[1:5, 1:5], 3)
```

Note that the entries of this matrix are *not* symmetric, like in the standard Breiger projection. For instance the weight of the link going form Evelyn to Laura is `r round(P.pp[1, 2], 3)`, but the weight of the link going from Laura to Evenly is  `r round(P.pp[2, 1], 3)`.

We can check this formally using the base `R` function `isSymmetric`:

```{r}
   isSymmetric(P)
   isSymmetric(P.pp)
```

Which returns `TRUE` for the usual Breiger projection but `FALSE` for the degree-weighted projection.

We can create weighted directed `igraph` objects from the degree-weighted projections using the function `graph_from_adjacency_matrix` as follows:

```{r}
   G.dp <- graph_from_adjacency_matrix(P.pp, mode = "directed", weighted = TRUE)
   G.dg <- graph_from_adjacency_matrix(P.gg, mode = "directed", weighted = TRUE)
```

Note that we specify the `mode` argument to be `directed` and the `weighted` argument to be `TRUE`.

We can of course plot the directed weighted graphs on each mode as usual:

```{r}
#| fig-height: 12
#| fig-width: 12
#| fig-cap: "Degree-weighted one mode projections."
#| label: fig-dgproj
#| fig-subcap: 
#|   - "Persons"
#|   - "Groups"
#| layout-ncol: 1

   set.seed(123)
   plot(G.dp, 
     vertex.size=8, vertex.frame.color="lightgray", 
     vertex.label.dist=2, edge.curved=0.2, 
     vertex.label.cex = 1.5, edge.color = "lightgray",
     edge.width = E(G.p)$weight*0.5)

   set.seed(123)
   plot(G.dg, 
     vertex.size=8, vertex.frame.color="lightgray", 
     vertex.label.dist=2, edge.curved=0.2, 
     vertex.label.cex = 1.5, edge.color = "lightgray",
     edge.width = E(G.g)$weight*0.5)
```

