---
title: "Community Detection in Two Mode Networks"
execute: 
  eval: true
  echo: true
  output: true
  warning: false
  message: false
format: 
   html:
      code-line-numbers: true
---

In the CA handout, we took a first stab at finding communities in two networks taking the simplest case of splitting a two-mode network into two-communities. We later saw, when discussing spectral clustering approaches approaches to community detection how to generalize that approach to split a two-mode network into multiple communities, while evaluating the quality of a given partition using the bipartite generalization of the modularity due to @barber07.

One limitation of Barber's approach is that it can only be used under the assumption that the number of communities is the *same* for both persons and groups. However, it could be that the optimal partition is actually one in which the people node set is split into a different number of clusters than the group node set.

So we need a way to evaluate the modularity of a partition when we have different number of communities on the people and group side. 

Let's load some data and see how it works. 

```{r}
   library(igraph)
   library(networkdata)
   g <- southern_women #southern women data
   A <- as.matrix(as_biadjacency_matrix(g)) #bi-adjacency matrix
   B <- as.matrix(as_adjacency_matrix(g)) #bipartite adjacency matrix
```

## The Dual Projection Approach to Bipartite Commmunity Detection

To estimate distinct number of communities across the separate modes we follow the dual projection approach of @everett_borgatti13.

We then perform the spectral decomposition of the bi-adjacency matrix, using Correspondence Analysis:

```{r}
   library(FactoMineR)
   ca.res <- CA(A, graph = FALSE, ncp = 10)
```

We can now use our trusty `k.cuts` function from the spectral community detection handout to come up with a bunch of partitions of people and groups based on the first three eigenvectors of the spectral decomposition of the modularity matrix:

```{r}
   k.cuts <- function(x, max = 9) {
      clus <- list()
      for (i in 1:max) {
         set.seed(456) #setting seed because kmeans uses random starting nodes for cluster centroids
         k <- i + 1
         clus[[i]] <- kmeans(x, k)$cluster
         }
      return(clus)
   }
   k.cuts.p <- k.cuts(ca.res$row$coord[, 1:6])
   k.cuts.g <- k.cuts(ca.res$col$coord[, 1:6])
```

## Finding the Bi-partition with the Best Modularity

OK, now that we have some candidate communities for both the people and groups (stored in the `k.cuts.p` and `k.cuts.g` objects) we need to come up with a way of measuring the modularity for a candidate partition that puts the people into $m$ and the groups into $n$ communities. 

As an example, let's pick the solution that partitions people into four and groups into three communities:

```{r}
   C.p <- k.cuts.p[[3]]
   C.g <- k.cuts.g[[2]]
```

Given this information we can create a reduced version of the bipartite adjacency matrix counting the proportion of edges that go from person community $i$ to group community $j$:

```{r}
   d <- max(C.p, C.g)
   e <- matrix(0, d, d)
   for (i in 1:d) {
      for (j in 1:d) {
         e[i, j] = sum(A[C.p == i, C.g == j])
      }
   }
   e
```

@suzuki_wakita09 based on work by @murata09 suggest  using the `e` matrix above to compute the modularity of this person/group assignment according to the following formula:

$$
Q = \sum_{i, j} = \frac{e_{ij}}{e_{i+}}\left(\frac{e_{ij}}{vol(A)} - \frac{e_{i+}e_{j+}}{vol(A)^2}\right)
$$

Where $e_{ij}$ is the fraction of edges connecting people in the $i^{th}$ person-community to groups in the $j^{th}$ group-community, $e_{i+}$ is the fraction of edges originating from person-community $i$ (the corresponding row sum of `e`), $e_{j+}$ is the fraction of edges originating from nodes in community $j$ (the corresponding column sum of `e`), and $vol(A)$ is the sum of the entries in the bi-adjacency matrix. 

So the idea is that given a partition of the person nodes into `m` communities and a partition of the group nodes into `n` communities, we can generate an `e` matrix like the one above and compute the corresponding modularity of that person/group partition using the above equation.

Let's see how this would work.

To do this, first let's package the steps we used to find calculate `e` from a partition and to find $Q$ from the Suzuki-Wakita modularity equation into a function:

```{r}
   find.mod <- function(w, x, y) {
      a <- max(c(x, y))
      b <- matrix(0, a, a)
      for (i in 1:a) {
         for (j in 1:a) {
            b[i, j] = sum(w[x == i, y == j]) 
         }
      }
      Q <- 0
      c <- rowSums(b)
      d <- colSums(b)
      e <- sum(w)
      for (i in 1:a) {
         for (j in 1:a) {
            if (c[i] > 0) {
               Q <- Q + (b[i, j]/c[i] * (b[i, j]/e - (c[i]*d[j])/e^2))
            }
         }
      }
   return(Q)
   }
```

So for the `e` matrix from the above example, $Q$ would be:

```{r}
   find.mod(A,  k.cuts.p[[3]],  k.cuts.g[[2]])
```

Which looks like a positive number. But is it the biggest of all the possible community partition combinations between people and groups?

To answer this question, we can use the `find.mod` function to compute the modularity between *every pair* of partition combinations between people and groups that we calculated earlier; since we computed eight different partitions for people and groups this leads to $8 \times 8 = 64$ pairs.

Here's a wrapper function over `find.mod` that computes the corresponding modularity values for each pair:

```{r}
   mod.mat <- function(d) {
     q <- matrix(0, d, d)
      for (i in 1:d) {
         for (j in 1:d) {
               q[i, j] <- find.mod(A,  k.cuts.p[[i]],  k.cuts.g[[j]])
         }
      }
    return(q)
   }
   Q <- round(mod.mat(8), 3)
   Q
```

Interestingly, the analysis suggests that the maximum modularity $Q = `r Q[4, 1]`$ is obtained with a partition of people into five communities and groups into two communities corresponding to cells $(4, 1)$ of the above matrix.

Here's what this community assignment looks like in the Southern Women data:

```{r}  
#| fig-height: 12
#| fig-width: 12
#| fig-cap: "Spectral Clustering of Nodes in the Southern Women Data with Optimal Community Assignment Obtained via the Suzuki Modularity, with Five Communities for People, Two Communities for Groups."
#| echo: false

   V(g)$type <- bipartite_mapping(g)$type
   V(g)$shape <- ifelse(V(g)$type, "square", "circle")
   V(g)$color <- c(k.cuts.p[[4]], k.cuts.g[[1]] + 5)
   set.seed(456)
   plot(g, 
     vertex.size=7, vertex.frame.color="lightgray", 
     vertex.label.dist = 1.5, edge.curved=0.2, 
     vertex.label.cex = 1.15)
```

The analysis separates two groups of densely connected actors of size six and five, respectively, namely, $\{Brenda, Theresa, Laura, Frances, Evelyn, Eleanor\}$ and $\{Katherine, Nora, Sylvia, Myrna, Helen\}$ along with their corresponding events from the one another. In the same way, $\{Pearl, Dorothy, Ruth, Verne\}$ form a community of more peripheral actors who selectively attend the more popular events; $\{Flora, Olivia\}$ are a two-actor community occupying the most peripheral position. Among the core set of actor $\{Charlotte\}$ occupies a singleton-community of her own. 

Events are partitioned into two broad groups: One the one hand, we have those selectively selectively attended by the larger community of densely connected actors along with the most popular events; on the other hand, we have the events selectively attended by the smaller group of densely connected actors. 






